Language Skills in an Ethnically Segmented Labour market: Estonia 1989 -- 2012

### 2018-07-07 paper/ijm.tex
methos overwritten

The Estonian economy, a part of the Soviet economy up to 1991, successfully
re-oriented to the Western markets and has been growing rapidly since
the mid-1990s.  
  As
a bi-product of the re-adjustment, English skills become extremely
valuable while the supply was still rather limited during the early
years of transition.

@@

The table indicates that there are substantial disparities between
Estonians and Russians in wage and unemployment. For instance, the
average monthly salary\footnote{When interpreting the wage,
  one should keep in mind that during the period
	of observation, 2000-2012, the average salary increased by 180\%,
	from \EUR{314} to \EUR{887} monthly.} of Estonian
males who are not fluent in English is \EUR{532} while for Russians it
is only \EUR{452}. However, we also see that Russians who speak
English earn approximately 40\% more than those who do not (\EUR{452}
versus \EUR{326} monthly for men and \EUR{307} versus \EUR{221} for
women). Estonian skills are associated with lower income premium, 15\%
for men (\EUR{378} versus \EUR{329} monthly) and 30\% for women
(\EUR{260} versus \EUR{203}).  Similarly, English is more closely
associated with income for Estonians than Russian language. In
contrast, unemployment shows a different picture. For ethnic Russians,
it is Estonian language fluency that is more closely associated with
low unemployment, those fluent in Estonian have 11.2 percent
unemployment rate while for those not fluent it is 17.3 percent.  The
corresponding numbers for English are 11.2 and 15.1.  For Estonians,
English is more closely associated with unemployment than Russian:
Estonians fluent in English have in average 5.0 percent unemployment
while those who are not fluent have 8.8 percent.  The difference
related to Russian fluency is only 1.6 percentage points.  However,
these associations are not easy to interpret as English skills are
also related to education, metropolitan residence, and general
ability.

% Table generated by Excel2LaTeX from sheet 'subgroups_by_ethnic_language'
\begin{sidewaystable}[htbp]
	\centering
	\caption{Averages and relative frequencies of selected explanatory
		variables 2000 -- 2012}
	\begin{tabular}{l|rrrr|rrrr}
		\toprule
		&  \multicolumn{4}{c|}{Estonians}   &   \multicolumn{4}{c}{Russians}    \\ \midrule
		Variable                    & RUS 0  & RUS 1  & ENG 0  & ENG 1  & EST 0  & EST 1  & ENG 0  & ENG 1  \\ \midrule
		Wage, men (EUR)             & 543.86 & 537.22 & 438.53 & 738.63 & 401.59 & 489.96 & 388.91 & 660.39 \\
		Wage, women (EUR)           & 345.51 & 376.98 & 301.13 & 488.83 & 238.25 & 338.62 & 266.87 & 435.60 \\
		Unemployment (\%)           & 8.74   & 7.13   & 8.83   & 4.99   & 17.30  & 11.18  & 15.06  & 11.23  \\
		EST 1 (\%)                  & 99.97  & 99.58  & 99.50  & 99.95  & 0.00   & 100.00 & 40.84  & 80.97  \\
		RUS 1 (\%)                  & 0.00   & 100.00 & 77.35  & 82.86  & 99.90  & 99.04  & 99.65  & 98.77  \\
		ENG 1 (\%)                  & 29.47  & 37.16  & 0.00   & 100.00 & 6.70   & 30.69  & 0.00   & 100.00 \\
		College Degree (\%)         & 15.61  & 23.44  & 10.77  & 41.38  & 12.66  & 28.64  & 13.93  & 49.16  \\
		Married (\%)                & 40.09  & 55.71  & 54.11  & 49.53  & 63.54  & 60.78  & 63.38  & 56.99  \\
		Inter-ethnic household (\%) & 1.34   & 5.20   & 4.79   & 3.70   & 5.07   & 19.46  & 11.97  & 12.17  \\
		Age                         & 37.05  & 41.83  & 42.39  & 38.05  & 42.14  & 40.54  & 42.15  & 37.83  \\
		Kids (number)               & 1.08   & 1.00   & 1.00   & 1.00   & 0.71   & 0.76   & 0.75   & 0.68   \\
		\textbf{Residence County }  &        &        &        &        &        &        &        &        \\
		Metropolitan (\%)           & 17.74  & 22.58  & 14.03  & 35.25  & 35.43  & 56.37  & 38.82  & 75.50  \\
		North-East (\%)             & 0.64   & 3.14   & 3.46   & 1.10   & 53.84  & 15.85  & 40.90  & 11.54  \\
		\textbf{Workplace}          &        &        &        &        &        &        &        &        \\
		Tenure (years)              & 6.17   & 7.33   & 7.45   & 6.40   & 8.19   & 7.23   & 8.08   & 5.99   \\
		Public sector (\%)          & 20.03  & 24.81  & 21.79  & 27.55  & 11.62  & 21.67  & 16.28  & 18.14  \\
		Primary sector (\%)         & 14.10  & 9.80   & 13.00  & 6.43   & 9.91   & 7.46   & 8.82   & 8.11   \\
		Secondary sector (\%)       & 31.02  & 27.20  & 32.63  & 19.41  & 47.21  & 29.72  & 41.54  & 24.07  \\
		Tertiary sector (\%)        & 26.17  & 30.12  & 26.52  & 34.46  & 27.17  & 33.96  & 28.37  & 40.96  \\
		\# Observations             & 14,943 & 57,270 & 46,525 & 25,688 & 13,805 & 12,830 & 21,773 & 4,862  \\ \bottomrule
	\end{tabular}%
	\label{tab:descriptive}%
	\caption*{\small
		Workplace--related variables are reported for employed individuals,
		other variables for the complete workforce.  Sample is limited to persons
		25-55 year old.\\
		\emph{RUS 0} refers to individuals who cannot speak Russian and
		\emph{RUS 1} to those who can. Analogously for English and
		Estonian.\\
		Primary sector = Agriculture,
		Fishing, Mining \\ Secondary sector = Manufacturing,
		Construction, Electricity \\ Tertiary (service) sector =
		Wholesale, Hotels, Transport, Financial intermediation, Real
		estate \\ Public sector = Public administration, Education,
		Health}
\end{sidewaystable}%

@@

Average age reflects the trend of younger generations
being more fluent in English and Estonian while older Estonians being
better in Russian. The family variables follow the
respective age distribution. Not surprisingly, inter-ethnic
households include disproportionally large share of Russian speaking
Estonians and Estonian speaking Russians.
The average experience follows the same pattern as age.
Public sector jobs are associated with disproportionally large share of
Estonian fluency among Russian workers, probably reflecting the formal
language requirements and little room for ethnic segregation in that
sector.

@@

We analyse the relationship between labour market outcomes and language
skills by ordinary multivariate regression models.
Several related studies use instrumental variables to correct for
omitted variable bias and measurement errors \citep{Chiswick1995, bleakley+chin2004}.
%\ot{more examples here}
We refrain from doing that. Our decision is mainly data driven,
there are no suitable instruments in our data,\footnote{We run an IV
	regression using age--at--immigration to instrument Estonian skills
	for those Russians who were in the country at 1990.  The results are
	not statistically significant at any conventional level.} but it also facilitates
comparison with other studies.For instance,
\citet{azam+2013EDandCC} and \citet{paolo+tansel2015JofDevStud} only report OLS results.

There is also evidence from job language requirements suggesting that
the reduced-form evidence is not too different from the causal effect
\citep{beblavy+2016CEPS}.

\citet{azam+2013EDandCC} and \citet{paolo+tansel2015JofDevStud}, who use proxy variables to control for
individual ability, find these having little influence on the effect
of language skills.

We estimate unemployment probability by linear probability model
(LPM) to avoid unnecessary parametric
assumptions in the corresponding binary choice models.\footnote{The
	results of corresponding probit models are virtually equal to the LPM results.}
Similarly, we estimate the effect of language skills on wage by
OLS. We correct the estimates for the panel data structure by
clustering the standard errors by individuals.

%\fxnote{Add here another index g to denote the group? I used it in other articles, because it was requested and it makes sense}
We estimate the basic model in the following form:
\begin{equation}
	\label{eq:specification}
	\begin{split}
		y_{it} ^{g} = &\: \alpha^{g}{}' X_{it}^{g} + \vec{\beta}^{g}{}' \cdot \vec{L}_{it}^{g} +
		\eta_{i}^{g} +
		\upsilon_{t}^{g} + \epsilon_{it}^{g},
	\end{split}
\end{equation}
where $y_{it}^{g}$ is the outcome (either unemployment indicator
or logarithm of net monthly wage)
of individual $i$ in wave $t$ from ethnic group g; $\vec{X}^{g}$ are the other individual
characteristics for both groups; $L$ is the vector of
language skill descriptors, $eta_{i}^{g}$ is the individual-specific effect, $\upsilon_{t}^{g}$ are the year dummies and $\epsilon_{it}^{g}$
is the idiosyncratic error term. We denote the groups g by E (Estonian) and R (Russian).
As we use panel data, we cluster the
standard errors on individuals.  When analysing the time trend, we
replace the main language effects in~\eqref{eq:specification} by the
\fxnote*{Did we estimate it really that way? At least not in my code. I found the code for estimations with these interactions, but I don't remember that I used these interactions for the tables.}{corresponding interactions}
\begin{align}
	\label{eq:time_interaction}
	\beta_{RUS,t} \cdot RUS_{it} \cdot \upsilon_{t} +
	\beta_{EST,t} \cdot EST_{it} \cdot \upsilon_{t} +
	\beta_{ENG,t} \cdot ENG_{it} \cdot \upsilon_{t}.
\end{align}

We estimate three models that differ in terms of the number of
control variables. The first one only includes the language skills
and year dummies. The second one also includes individual
characteristics (age, education) and family descriptors (marital status, having kids up to 17 years old) and county of residence.
Finally, the third model additionally includes industry, occupation, and other
workplace descriptors.
\Cref{tab:models} summarises this description of the models.
%SKB: This table should have been in the article at some point, but I must have removed it. I use in my Quantile Regression article
\begin{table}[hbtp]
	\centering
	\caption{Groups of explanatory variables included in different models}
	\label{tab:models}
	\begin{tabular}{r|ccc}
		\toprule
		& Model 1 & Model 2 & Model 3 \\ \midrule
		year dummies       & \V      & \V      & \V      \\
		indiv. charact.    &         & \V      & \V      \\
		workplace charact. &         &         & \V      \\ 
		\bottomrule
	\end{tabular}
\end{table}

>>>>

The Estonian economy, a part of the Soviet economy up to 1991, successfully
re-oriented to the Western markets and has been growing rapidly since
the mid-1990s.  
\fxwarning{ot: a few numbers about economic growth would be good}
  As
a bi-product of the re-adjustment, English skills become extremely
valuable while the supply was still rather limited during the early
years of transition.

@@

The table indicates that there are substantial disparities between
Estonians and Russians in wage and unemployment. For instance, the
average monthly salary\footnote{When interpreting these figures,
  one should keep in mind that during this period, 2000-2012, the average monthly salary grow by 180\%,
  from \EUR{314} to \EUR{887} while the language knowledge changed rapidly as
  well.  However, the ethnic composition of the workforce remained
  roughly constant.} of Estonian
males who are not fluent in English is \EUR{532} while for Russians it
is only \EUR{452}. However, we also see that Russians who speak
English earn approximately 40\% more than those who do not (\EUR{452}
versus \EUR{326} monthly for men and \EUR{307} versus \EUR{221} for
women). Estonian skills are associated with lower income premium, 15\%
for men (\EUR{378} versus \EUR{329} monthly) and 30\% for women
(\EUR{260} versus \EUR{203}).  Analogously, English is more closely
associated with income for Estonians compared to Russian language. In
contrast, unemployment shows a different picture. For ethnic Russians,
it is Estonian language fluency that is more closely associated with
low unemployment, those fluent in Estonian have 11.2 percent
unemployment rate while for those not fluent it is 17.3 percent.  The
corresponding numbers for English are 11.2 and 15.1.  For Estonians,
English is more closely associated with unemployment than Russian:
Estonians fluent in English have in average 5.0 percent unemployment
while those who are not fluent have 8.8 percent.  The difference
related to Russian fluency is only 1.6 percentage points.  However,
these correlations are not easy to interpret as besides of the strong
time trends, English skills are
also related to education, metropolitan residence, and general
ability.

% Table generated by Excel2LaTeX from sheet 'subgroups_by_ethnic_language'
\begin{sidewaystable}[tbp]
	\centering
	\caption{Averages and relative frequencies of selected explanatory
		variables 2000 -- 2012}
	\begin{tabular}{l|rrrr|rrrr}
		\toprule
		&  \multicolumn{4}{c|}{Estonians}   &   \multicolumn{4}{c}{Russians}    \\ \midrule
		Variable                    & RUS 0  & RUS 1  & ENG 0  & ENG 1  & EST 0  & EST 1  & ENG 0  & ENG 1  \\ \midrule
		Wage, men (EUR)             & 543.86 & 537.22 & 438.53 & 738.63 & 401.59 & 489.96 & 388.91 & 660.39 \\
		Wage, women (EUR)           & 345.51 & 376.98 & 301.13 & 488.83 & 238.25 & 338.62 & 266.87 & 435.60 \\
		Unemployment (\%)           & 8.74   & 7.13   & 8.83   & 4.99   & 17.30  & 11.18  & 15.06  & 11.23  \\
		EST 1 (\%)                  & 99.97  & 99.58  & 99.50  & 99.95  & 0.00   & 100.00 & 40.84  & 80.97  \\
		RUS 1 (\%)                  & 0.00   & 100.00 & 77.35  & 82.86  & 99.90  & 99.04  & 99.65  & 98.77  \\
		ENG 1 (\%)                  & 29.47  & 37.16  & 0.00   & 100.00 & 6.70   & 30.69  & 0.00   & 100.00 \\
		College Degree (\%)         & 15.61  & 23.44  & 10.77  & 41.38  & 12.66  & 28.64  & 13.93  & 49.16  \\
		Married (\%)                & 40.09  & 55.71  & 54.11  & 49.53  & 63.54  & 60.78  & 63.38  & 56.99  \\
		Inter-ethnic household (\%) & 1.34   & 5.20   & 4.79   & 3.70   & 5.07   & 19.46  & 11.97  & 12.17  \\
		Age                         & 37.05  & 41.83  & 42.39  & 38.05  & 42.14  & 40.54  & 42.15  & 37.83  \\
		Kids (number)               & 1.08   & 1.00   & 1.00   & 1.00   & 0.71   & 0.76   & 0.75   & 0.68   \\
		\textbf{Residence County }  &        &        &        &        &        &        &        &        \\
		Metropolitan (\%)           & 17.74  & 22.58  & 14.03  & 35.25  & 35.43  & 56.37  & 38.82  & 75.50  \\
		North-East (\%)             & 0.64   & 3.14   & 3.46   & 1.10   & 53.84  & 15.85  & 40.90  & 11.54  \\
		\textbf{Workplace}          &        &        &        &        &        &        &        &        \\
		Tenure (years)              & 6.17   & 7.33   & 7.45   & 6.40   & 8.19   & 7.23   & 8.08   & 5.99   \\
		Public sector (\%)          & 20.03  & 24.81  & 21.79  & 27.55  & 11.62  & 21.67  & 16.28  & 18.14  \\
		Primary sector (\%)         & 14.10  & 9.80   & 13.00  & 6.43   & 9.91   & 7.46   & 8.82   & 8.11   \\
		Secondary sector (\%)       & 31.02  & 27.20  & 32.63  & 19.41  & 47.21  & 29.72  & 41.54  & 24.07  \\
		Tertiary sector (\%)        & 26.17  & 30.12  & 26.52  & 34.46  & 27.17  & 33.96  & 28.37  & 40.96  \\
		\# Observations             & 14,943 & 57,270 & 46,525 & 25,688 & 13,805 & 12,830 & 21,773 & 4,862  \\ \bottomrule
	\end{tabular}%
	\label{tab:descriptive}%
	\caption*{\small
		Workplace--related variables are reported for employed individuals,
		other variables for the complete workforce.  Sample is limited to persons
		25-55 year old.\\
		\emph{RUS 0} refers to individuals who cannot speak Russian and
		\emph{RUS 1} to those who can. Analogously for English
                (ENG) and
		Estonian (EST).\\
		Primary sector = Agriculture,
		Fishing, Mining \\ Secondary sector = Manufacturing,
		Construction, Electricity \\ Tertiary (service) sector =
		Wholesale, Hotels, Transport, Financial intermediation, Real
		estate \\ Public sector = Public administration, Education,
		Health}
\end{sidewaystable}%

@@

Average age reflects the trend of younger generations
being more fluent in English and Estonian while older Estonians being
better in Russian. The family status--related variables correspond to the
respective age distribution. Not surprisingly, inter-ethnic
households include disproportionally large share of Russian speaking
Estonians and Estonian speaking Russians.
The average experience follows the same pattern as age.
Public sector jobs are associated with disproportionally large share of
Estonian fluency among Russian workers, probably reflecting the formal
language requirements and little room for ethnic segregation in that
sector.

@@

We analyse the relationship between labour market outcomes and language
skills by ordinary multivariate regression models.
Several related studies use instrumental variables to correct for
omitted variable bias and measurement errors \citep{Chiswick1995, bleakley+chin2004}.
We refrain from doing that. Our decision is mainly data driven,
there are no suitable instruments in our data,\footnote{We run an IV
	regression using age--at--immigration to instrument Estonian skills
	for those Russians who were in the country at 1990.  The results are
	not statistically significant at any conventional level.} but it also facilitates
comparison with other studies.  For instance,
\citet{azam+2013EDandCC} and \citet{paolo+tansel2015JofDevStud} only
report OLS results, and, even more, find that the available proxy
variables have little influence on the final results.

We use linear probability models (LPM) to estimate the relationship between
language skills and unemployment, mainly in order to avoid unnecessary parametric
assumptions in the corresponding binary choice models.\footnote{The
	results of corresponding probit models are virtually equal to the LPM results.}
Similarly, we estimate the effect of language skills on wage by
OLS.

We estimate the basic model in the following form:
\begin{equation}
	\label{eq:specification}
	\begin{split}
		y_{it} ^{g} = &\: \alpha^{g}{}' \vec{X}_{it}^{g} + \vec{\beta}^{g}{}' \cdot \vec{L}_{it}^{g} +
		\eta_{i}^{g} +
		\upsilon_{t}^{g} + \epsilon_{it}^{g},
	\end{split}
\end{equation}
where $y_{it}^{g}$ is the outcome (either unemployment indicator
or logarithm of net monthly wage)
of individual $i$ in year $t$ from ethnic group $g$ (either Estonian
or Russian); $\vec{X}$ are the other individual
characteristics; $\vec{L}$ is the vector of
language skill descriptors, $\eta_{i}^{g}$ is the individual-specific effect, $\upsilon_{t}^{g}$ are the year dummies and $\epsilon_{it}^{g}$
is the idiosyncratic error term. 
\fxerror{ot: what's eta?  do you do random effects?}
As we use panel data, we cluster the
standard errors on individuals.  When analysing the time trend, we
replace the main language effects in~\eqref{eq:specification} by the
corresponding interactions
\fxwarning{ot: ensure this is what we do :-)}
\begin{align}
	\label{eq:time_interaction}
	\beta_{RUS,t} \cdot RUS_{it} \cdot \upsilon_{t} +
	\beta_{EST,t} \cdot EST_{it} \cdot \upsilon_{t} +
	\beta_{ENG,t} \cdot ENG_{it} \cdot \upsilon_{t}.
\end{align}

We estimate three models that differ in terms of the number of
control variables. The first one only includes the language skills
and year dummies. The second one also includes individual
characteristics (age, education) and family descriptors (marital status, having kids up to 17 years old) and county of residence.
Finally, the third model also adds industry, occupation, and other
workplace descriptors.  Note that in case of unemployment probability,
we only estimate the first two models.
